---
title: "DATA612_Project 5 | IMPLEMENTING A RECOMMENDER SYSTEM ON SPARK"
author: "Amber Ferger, Charlie Rosemond, Juanelle Marks" 
date: "6/14/2020"
output: 
    html_document:
     theme: lumen
     highlight: tango
     toc: TRUE
---
## Assignment

The goal of this project is give you practice beginning to work with a distributed recommender system. It is sufficient for this assignment to build out your application on a single node.

Adapt one of your recommendation systems to work with Apache Spark and compare the performance with your previous iteration. Consider the efficiency of the system and the added complexity of using Spark. You may complete the assignment using PySpark (Python), SparkR (R) , sparklyr (R), or Scala.

Please include in your conclusion: For your given recommender system’s data, algorithm(s), and (envisioned) implementation, at what point would you see moving to a distributed platform such as Spark becoming necessary?

You may work on any platform of your choosing, including Databricks Community Edition or in local mode. You are encouraged but not required to work in a small group on this project.


## Motivation

In this assignment, we create two alternating least squares (ALS) recommenders--one with *recommenderlab* and one with Apache Spark via *sparklyr*--and compare their performance. Our goal is to gain experience with Spark and begin to assess its advantages over a non-distributed platform.


## Data source

Our data source is the [Jester](http://eigentaste.berkeley.edu/dataset/) dataset, which contains data from 24,983 users who have rated 36 or more jokes. Ratings are real values ranging from -10.00 to +10.00 (the value "99" corresponds to "null" = "not rated"). Each row in the dataset represents a different user, and the first column represents the total number of jokes the individual has rated. The remaining 100 columns give the ratings for each joke. 


### Load libraries

```{r message=FALSE, include=FALSE}
# load libraries
library(tidyverse)
library(pander)
library(knitr)
library(dplyr)
library(recommenderlab)
library(readxl)
```

### Read and prepare data

We start by reading in the jester data, replacing the current null label ("99") with NULL values, and creating

```{r, warning=FALSE, message=FALSE}
# Read jester data
jester <- data.frame(read_xls("jester-data-1.xls", col_names = FALSE))
colnames(jester) <- c("ratingCount", seq(100))
row.names(jester) <- 1:nrow(jester)

# Remove num jokes column
ratings <- jester[-1]

# Replace 99 (no rating) with NULL
ratings[ratings == 99] <- NA

# Subset to first 5000 users
ratings <- ratings[1:5000,]

# Create large dgCMatrix
ratings <- as.matrix(ratings)
finalRatings <- as(ratings, 'realRatingMatrix')
```

## Exploratory data analysis

We create two visualizations that describe the data, including histograms depicting joke ratings per user and ratings per joke. We also compare the mean and median joke ratings as well as the overall distribution of ratings.

Approximately 1500 users rated between 95 and 100 jokes each, and approximately 1000 users rated between 70 and 75 jokes each. 

```{r}
jokeCount <- rowCounts(finalRatings)
hist(jokeCount,
     main = 'Number of Jokes Rated per User',
     xlab = 'Number of Jokes Rated',
     ylab = 'Number of Users')
```


Approximately 35 jokes have 4500 and 5000 ratings each, meaning nearly every user rated those jokes.

```{r}
ratingCount <- colCounts(finalRatings)
hist(ratingCount,
     main = 'Number of Users Rating each Joke',
     xlab = 'Number of Users that Rated Joke',
     ylab = 'Number of Jokes')
```
  

The mean rating is approximately 0.89, or just above neutral. Relative to the mean, the median of 1.5 indicates a slight left skew--a shape borne out by a histogram of the ratings.

```{r}
mean(finalRatings@data@x, na.rm = T)
summary(finalRatings@data@x, na.rm = T)
hist(finalRatings@data@x, main = "Histogram of ratings")
```

## Recommenders

We compare alternating least squares recommenders built using *recommenderlab* and Spark via *sparklyr*.

### recommenderlab

## Discussion

## Conclusion

## Reference
